\textbf{COVID-19 vaccination in Ontario: Exploring intra-provincial variations within Health Regions and socio-economic strata}

### 2023-03-22 manuscript_v2/main.tex
updated analyses, removed shp files

\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{setspace}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
  \setmainfont[]{Palatino Linotype}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\begin{document}
\maketitle
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, enhanced, breakable, interior hidden, sharp corners, frame hidden]}{\end{tcolorbox}}\fi

@@

The COVID-19 pandemic continues to be a worldwide public health concern.
Although vaccines against this disease were rapidly developed,
vaccination uptake has not ben equal across all the segments of the
population. In particular, it has been shown that there have been
differences in vaccine uptake across different segments of the
population. However, there are also differences in vaccination across
geographical areas, which might be important to consider in the
development of future public health policies against COVID-19. In this
study, we examined the relationship between vaccination status (having
received the first dose of a COVID-19 vaccine), and different
socio-economic and geographical factors. Our results show differences in
vaccination due to race/ethnicity, income, Health Regions (geographical
areas used for health service access in Ontario), and their
interactions. In particular, we show that individuals who identified as
Arab/Middle Eastern, Black, or Latin American, had significantly lower
odds of vaccination than White/Caucasian individuals (ORs=0.31, 0.32,
0.28, and \emph{p}=0.004, \emph{p}\textless0.001 and \emph{p}=0.004,
respectively), and that individuals with a household income below CAD
25,000 who identified as Arab/Middle Eastern (OR=3.05, \emph{p}=0.013),
Black (OR=3.19, \emph{p}=0.004), Latin American (OR=2.80,
\emph{p}=0.041), or that belonged to other minority groups (OR=4.59,
\emph{p}\textless0.001) had higher odds of vaccination than individuals
from the same racial/ethnic group in higher income brackets. Finally, we
also identified lower odds of vaccination within different minority
groups in West Health Region. This study shows that there is an ongoing
need to better understand and address differences in vaccination uptake
across diverse segments of the population that have been largely
impacted by the pandemic.

@@

The COVID-19 pandemic continues around the world with more than 600
million confirmed cases as of November of
2022\textsuperscript{\protect\hyperlink{ref-WHO-Covid}{1}}. During the
first months of the pandemic in early 2020, non-pharmaceutical
interventions such as masking and social distancing were the only
methods available to manage the spread of the disease; however, the
rapid development of vaccines permitted their approval and use in some
countries towards the last month part of 2020. For example, in the US
and Canada vaccine campaigns began in mid-December of
2020\textsuperscript{\protect\hyperlink{ref-tanne2020}{2},\protect\hyperlink{ref-bogoch2022}{3}}.
Although it has been estimated that COVID-19 vaccines have prevented
around 14 millions of deaths
worldwide\textsuperscript{\protect\hyperlink{ref-watson2022}{4}}, the
rollout of COVID-19 vaccines has faced multiple challenges since its
inception.

Indeed, multiple obstacles that have complicated vaccination efforts
against COVID-19 have been identified: inequality in vaccine access,
vaccine hesitancy, and differences in vaccination rates across different
segments of the
population\textsuperscript{\protect\hyperlink{ref-gerretsen2021}{5}--\protect\hyperlink{ref-malik2020}{7}}.
In the case of Canada, lower vaccine uptake has been associated with
socio-economic factors such as younger age, educational level, presence
of children in the household, lack of a regular healthcare provider,
ethnic origin, and financial
instability\textsuperscript{\protect\hyperlink{ref-guay2022}{8}--\protect\hyperlink{ref-hussain2022}{10}}.

Additionally, vaccination is influenced by changes in geography. In this
regard, it has been shown that there have been spatial differences in
COVID-19 vaccination rates due to regional differences in attitudes
towards
vaccination\textsuperscript{\protect\hyperlink{ref-malik2020}{7}},
geographical differences in vaccine access and supply, vaccination
location availability, and lack of prioritization of vulnerable
groups\textsuperscript{\protect\hyperlink{ref-bogoch2022}{3},\protect\hyperlink{ref-nguyen2021}{11}}.

Studies that analyze geographical variations in vaccine uptake are
important as they can help inform public health decision-makers to
design policies to that consider spatial changes to address vaccination
disparities. In this regard, previous geographical (spatial) analyses of
vaccination rates have shown that variations in vaccine uptake can occur
within small governmental administrative units (e.g., counties in the
case of the
US)\textsuperscript{\protect\hyperlink{ref-mollalo2021}{12}--\protect\hyperlink{ref-bhuiyan2022}{15}},
and that geographical analyses can be predictive of booster uptake
patterns\textsuperscript{\protect\hyperlink{ref-wood2022}{16}}.

In Canada, studies that have used a spatial approach to analyze vaccine
uptake have shown disparities in vaccination rates across low and high
income neighborhoods in the city of
Toronto\textsuperscript{\protect\hyperlink{ref-choi2021}{17}}, among
adolescents from deprived neighborhoods in the city of
Montreal\textsuperscript{\protect\hyperlink{ref-mckinnon2021}{18}}, and
have also highlighted disparities in vaccination status depending on
age, income, and ethnic origin at the provincial
level\textsuperscript{\protect\hyperlink{ref-guay2022}{8}}. However,
there is limited information on differences in vaccination status inside
the provinces. Such analysis is important as it can help identify
inequalities that may exist within these geographical areas while
providing a granular view of intra-provincial differences that can help
understand the barriers for vaccine uptake.

In this study, we examined self-reported COVID-19 vaccination status
within the province of Ontario using a combination of socio-economic
factors(such as ethnic origin, age, and income) and geographical
analysis (at the level of the Health Regions) in order to determine
intra-provincial differences in vaccine uptake and address the ongoing
need of socio-economic information that can provide a rationale for the
disparities in vaccination observed within some racial
groups\textsuperscript{\protect\hyperlink{ref-cnat2022b}{19}}.

@@

\hypertarget{data-source-survey-overview}{%
\subsection{Data source: survey
overview}\label{data-source-survey-overview}}

We used data from the \emph{Survey of COVID-19 related Behaviours and
Attitudes}, a repeated cross sectional survey focused on the Canadian
province of Ontario which ran from Sept 30, 2021 until January 17, 2022
and that was comissioned by the Fields Institute for Research in
Mathematical Sciences (henceforth Fields) and the Mathematical Modelling
of COVID-19 Task Force. The survey was conducted by a third-party
service provider (RIWI Corp.), under ethical guidance from the
University of Toronto.

Briefly, the survey was deployed used random domain intercept
technology, where if users clicked on a registered but commercially
inactive web link or typed in a web address for a site that was dormant,
they had a random chance of that link being temporarily managed by the
company that administered the survey and instead of coming across a
notification about the status of the site(``this page does not exist''),
the survey was deployed to the
user\textsuperscript{\protect\hyperlink{ref-sargent2022}{20}}. Users
then decided whether to anonymously participate, and those that
participated were able to exit the survey at any time. After the survey
closed, regardless if it was complete or incomplete, access was denied
to any further users with the same internet protocol (IP) address and
the domain entry point rotated such that if a user were to attempt to
access the survey again, share the link, or enter via the same address
using an alternative IP address, the survey would not deploy. This
effectively meant that a user could participate only once in the survey.

Additionally, users who indicated they were under the age of 16 were
exited from the survey without creating a record, furthermore, these
users were unable to navigate back to the ``age select'' screen. The
personal identifier information from each user that participated in the
survey was automatically scrubbed and replaced by a unique ID.
Respondents were drawn exclusively from the province of Ontario, as per
their devices meta-data.

\hypertarget{sec-socio-demographic-factors}{%
\subsection{Survey responses}\label{sec-socio-demographic-factors}}

Socio-economic information selected from the survey answers included age
group and income brackets, and race/ethnicity. The levels of each
socio-economic factor used for analysis appear in
Table~\ref{tbl-covariates}.

\hypertarget{tbl-covariates}{}
\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.3014}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.6986}}@{}}
\caption{\label{tbl-covariates}Selected socio-economic factors from the
survey}\tabularnewline
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
Variable
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Levels
\end{minipage} \\
\midrule()
\endfirsthead
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
Variable
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Levels
\end{minipage} \\
\midrule()
\endhead
Age group & 16-34,35-54,55 and over \\
Income bracket (CAD) & under 25,000, 25,000-59,999, 60,000 and above \\
Race/ethnicity & Arab/Middle Eastern, Black, East Asian/Pacific
Islander, Indigenous, Latin American, Mixed, South Asian, White
Caucasian, Other \\
\bottomrule()
\end{longtable}

Furthermore, the information on vaccination status was provided by
survey participants who answered the question ``Have you received the
first dose of the COVID vaccine?'', with possible answers ``yes'' and
``no''.

\hypertarget{data-cleaning}{%
\subsection{Data cleaning}\label{data-cleaning}}

The original dataset contained 39,029 entries (where each entry
corresponded to a set of answers provided by a unique respondent).
Following a preliminary analysis, the dataset was cleaned in order to
only contain the socio-economic information provided in
Table~\ref{tbl-covariates} and vaccination status. The cleaning process
also included removing outliers that were identified during the
preliminary analyses, and processing the geographical information in the
survey (city where the survey was responded) in order to match each city
to its correspondent Health Region.

\hypertarget{sec-geographical-location}{%
\subsubsection{Geographical location}\label{sec-geographical-location}}

For each survey participant certain data was automatically captured,
including the nearest municipality (city). This resulted in a total of
578 different municipalities within the clean dataset. Due to our
interest in analyzing the differences between Health Regions, we
assigned the city of each entry to its correspondent Health Region
following a multi-step process. Briefly, we used Local Health Integrated
Networks (LHINs) to assign a Health Region to each entry in the survey.
LHINs were the geographical divisions for health used by Ontario before
the adoption of the Health Regions; because of the lack of a publicly
available list of all municipalities within each Health Region, we used
a dataset of long-term care homes and LHINs to match each city to LHIN,
followed by matching each LHIN to a Health Region following the
information provided on the Ontario Health Website, where the list of
LHINs and corresponding Health Regions is available. In the case of
municipalities that did not appear in the long-term dataset, we manually
searched each city in the LHINs websites in order to provide
geographical information. The raw dataset, clean dataset, details of the
data cleaning process, and the addition of Health Region and LHIN
information are described in detail in the GitHub repository for this
paper, which can be found at
\url{https://github.com/aimundo/Fields_COVID-19/}.

Following an assessment of the number of entries corresponding to each
Health Region in the final dataset, only 107 observations (4.3\% of the
total) corresponded to cities located in the North West and North East
Health Regions. Due to the low number of entries, we omitted these
Health Regions from further analyses. Therefore, the total number of
unique entries used for analysis was 3,551 which included the East,
Central, Toronto, and West Health Regions.

\hypertarget{corrections}{%
\subsection{Corrections}\label{corrections}}

We identified differences between the proportions of all the
socio-economic factors included in the analysis
(Table~\ref{tbl-covariates}) and the 2016 Canada Census data for
Ontario. Additionally, because the Census divisions do not match the
exact boundaries of the Health Regions, we also obtained population
estimates for each Health Region from the Ontario Health website in
order to correct for the population totals. We used an iterative
proportional fitting procedure
(\emph{raking})\textsuperscript{\protect\hyperlink{ref-deming1940}{21}}
to correct for socio-economic factors and Health Region populations
using the R \texttt{survey} package. Details about the correction can be
found in the Appendix.

@@

We used a logistic regression model to estimate the probability of
vaccination depending on the socio-economic factors described in
Section~\ref{sec-socio-demographic-factors}, the Health Regions from
Section~\ref{sec-geographical-location}, and the interactions between
Race and Health Region, and Race and income as previous studies have
shown that socio-economic factors and their interactions are significant
predictors of intent of vaccination and vaccination
status\textsuperscript{\protect\hyperlink{ref-nguyen2022}{22}--\protect\hyperlink{ref-cnat2022a}{24}}.
The model appears in Equation~\ref{eq-model3},

\begin{equation}\protect\hypertarget{eq-model3}{}{
\begin{aligned}
\log \left( \frac{p\textrm{(vac)}}{1-p\textrm{(vac)}} \right) = \beta_0+ \beta_{1}\textrm{(Age group)} +\beta_{2} \textrm{ Race} + \beta_3 \textrm{ Health Region} + \beta_4 \textrm{ Income}+\\ \beta_5\textrm{(Health Region} \times \textrm{Race)} + \beta_6 \textrm{ (Income} \times \textrm{Race)}
\end{aligned}
}\label{eq-model3}\end{equation}

Where \(p\textrm{(vac)}\) indicates the probability of having received
the first dose of a COVID-19 vaccine, \(\beta_0\) indicates the
population intercept, and \(\beta_1...\beta_6\) indicate the
coefficients for each term. The the model was fitted using the function
\texttt{svyglm} from the \texttt{survey} R package in order to
incorporate the correction in sampling probability obtained from raking.

All analyses were conducted in RStudio (2022.12.0 Build 353), using R
4.2.2 and the packages
\texttt{survey}\textsuperscript{\protect\hyperlink{ref-lumley2011}{25}},\texttt{tidyverse}\textsuperscript{\protect\hyperlink{ref-wickham2019}{26}},
and \texttt{quarto}\textsuperscript{\protect\hyperlink{ref-quarto}{27}}.

@@

Table~\ref{tbl-model} shows the results of the logistic regression on
vaccination status using socio-economic factors (age group, income,
race), geographical areas (Health Regions) and the interactions between
income and race and Health Region and race. The reference groups were
set as follows: 16 to 34 years (age group), White Caucasian (Race),
Toronto (Health Region), CAD 60,000 and over (Income). There were no
statistically significant differences in vaccination rates within the
age groups from the survey, but significant odds ratios were estimated
for other covariates. Within household income brackets, individuals with
an income under CAD 25,000 or between CAD 25,000-59,999 had
significantly lower odds of vaccination than those with an income above
CAD 60,000 (ORs=0.37 and 0.59, respectively). Within Race/Ethnicity,
individuals who identified as Arab/Middle Eastern, Black, or Latin
American, had significantly lower odds of vaccination than those in the
White/Caucasian group (ORs=0.31, 0.32, 0.28, and
\emph{p}=0.004,\textless0.001 and 0.004, respectively); additionally,
those individuals in the Other Race/Ethnicity group (a group that
included Southeast Asian, Filipino, West Asian, and Minorities Not
Identified Elsewhere) had even lower odds of vaccination than the other
minority groups (OR=0.22, \emph{p}\textless0.001). Regarding Health
Regions, individuals that reported living in the West Health Region
(which comprises the regions of Waterloo and Niagara, the counties of
Wellington, Essex, and Lambton, and the cities of Hamilton, Haldimand,
Brant, and Chatham-Kent) had significantly higher odds of vaccination
than those in the Health Region of Toronto (OR=1.55, \emph{p}=0.029).

@@

\hypertarget{tbl-model}{}
\begin{longtable}{lccc}
\caption{\label{tbl-model}Multiple Regression Analysis-Predictors of Vaccination Status }\tabularnewline

\toprule
\textbf{Characteristic} & \textbf{OR} & \textbf{95\% CI} & \textbf{p-value}\\
\midrule
\endfirsthead
\multicolumn{4}{@{}l}{\textit{(continued)}}\\
\toprule
\textbf{Characteristic} & \textbf{OR} & \textbf{95\% CI} & \textbf{p-value}\\
\midrule
\endhead

>>>>

\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{setspace}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\begin{document}
\maketitle
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, interior hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, breakable, sharp corners, frame hidden]}{\end{tcolorbox}}\fi

@@

The COVID-19 pandemic continues to be a worldwide public health concern.
Although vaccines against this disease were rapidly developed,
vaccination uptake has not ben equal across all the segments of the
population. In particular, it has been shown that there have been
differences in vaccine uptake across different segments of the
population. However, there are also differences in vaccination across
geographical areas, which might be important to consider in the
development of future public health policies against COVID-19. In this
study, we examined the relationship between vaccination status (having
received the first dose of a COVID-19 vaccine), and different
socio-economic and geographical factors. Our results show differences in
vaccination due to race/ethnicity, income, Health Regions (geographical
areas used for health service access in Ontario), and their
interactions. In particular, we show that individuals who identified as
Arab/Middle Eastern, Black, or Latin American, had significantly lower
odds of vaccination than White/Caucasian individuals (ORs=0.31, 0.32,
0.28, and \emph{p}=0.004, \emph{p}\textless0.001 and \emph{p}=0.004,
respectively), and that individuals with a household income below CAD
25,000 who identified as Arab/Middle Eastern (OR=3.05, \emph{p}=0.013),
Black (OR=3.19, \emph{p}=0.004), Latin American (OR=2.80,
\emph{p}=0.041), or that belonged to other minority groups (OR=4.59,
\emph{p}\textless0.001) had higher odds of vaccination than individuals
from the same racial/ethnic group in higher income brackets. Finally, we
also identified lower odds of vaccination within certain minority groups
in the West Health Region, which comprises the regions of Waterloo and
Niagara, the counties of Wellington, Essex and Lambton, and the cities
of Hamilton, Haldimand, Brant, and Chatham-Kent. This study shows that
there is an ongoing need to better understand and address differences in
vaccination uptake across diverse segments of the population that have
been largely impacted by the pandemic.

@@

The vaccines against COVID-19 have been considered a major achievement
of modern medicine as their rapid
development\textsuperscript{\protect\hyperlink{ref-davis2022}{1}}
allowed the start of broad vaccination campaigns towards the end of 2020
in certain countries such as the
US\textsuperscript{\protect\hyperlink{ref-tanne2020}{2}} and
Canada\textsuperscript{\protect\hyperlink{ref-bogoch2022}{3}}. This made
some believe that vaccines were destined to be a determinant factor in a
rapid ending of the
pandemic\textsuperscript{\protect\hyperlink{ref-thelancet2021}{4}}.
However, although it has been estimated that COVID-19 vaccines have
prevented around 14 million of deaths
worldwide\textsuperscript{\protect\hyperlink{ref-watson2022}{5}}, their
implementation has been far from being equal to that of the vaccines of
smallpox and polio, which were implemented on a global scale and that
were indeed crucial to control these
diseases\textsuperscript{\protect\hyperlink{ref-kayser2021}{6}}. In
fact, the rollout of COVID-19 vaccines has faced multiple challenges
since its inception which ultimately have hampered the goal of global
immunity.

This ongoing lack of a successful global rollout of the COVID-19
vaccines is a multifaceted issue resulting from, as among other things,
the development of new variants due to inadequate public health
measures\textsuperscript{\protect\hyperlink{ref-li2021}{7}}, inequality
in vaccine access between high- and low-income
countries\textsuperscript{\protect\hyperlink{ref-gerretsen2021}{8},\protect\hyperlink{ref-tamey2022}{9}},
vaccine
hesitancy\textsuperscript{\protect\hyperlink{ref-nafilyan2021}{10}}, and
differences in vaccination uptake across different segments of the
population\textsuperscript{\protect\hyperlink{ref-malik2020}{11}}. In
particular, it is well established that differences in vaccination
uptake have been present even in the case of countries that have had
ample access to vaccines since 2020 (such as the US, the UK, and
Canada), where lower vaccine uptake has been observed within racial
minorities (i.e., individuals that identify as Black, Asian, or
Indigenous), and in individuals within low income
brackets\textsuperscript{\protect\hyperlink{ref-willis2021}{12}--\protect\hyperlink{ref-khubchandani2021}{15}}.
Reasons given for lower vaccine uptake in these cases have included
medical mistrust due to systemic medical racism, mistrust in vaccines,
and the influence of conspiracy
theories\textsuperscript{\protect\hyperlink{ref-willis2021}{12},\protect\hyperlink{ref-stoler2021}{14},\protect\hyperlink{ref-bogart2021}{16}--\protect\hyperlink{ref-freeman2020}{18}}.
More specifically, in the case of Canada, lower vaccine uptake has been
observed in young individuals, those with a low educational level,
households with children, those without a regular healthcare provider,
individuals that identify as part of a visible minorities or Indigenous,
and those that are financially
unstable\textsuperscript{\protect\hyperlink{ref-guay2022}{19}--\protect\hyperlink{ref-hussain2022}{21}}.

However, it is important to consider that vaccination uptake can also be
influenced by geographical (spatial) factors. In this regard,
differences in COVID-19 vaccination rates have been associated with
varied regional attitudes towards
vaccination\textsuperscript{\protect\hyperlink{ref-malik2020}{11}},
spatial differences in vaccine access and supply, vaccination location
availability, and lack of prioritization of areas where vulnerable
groups
reside\textsuperscript{\protect\hyperlink{ref-bogoch2022}{3},\protect\hyperlink{ref-nguyen2021}{22}}.
Other studies have also shown heterogeneity in vaccine uptake within
small governmental administrative units such as
counties\textsuperscript{\protect\hyperlink{ref-mollalo2021}{23}--\protect\hyperlink{ref-bhuiyan2022}{26}},
and that and that accounting for geographical differences in vaccination
can help predict patterns of booster
uptake\textsuperscript{\protect\hyperlink{ref-wood2022}{27}}. Overall,
the existing body of research on geographical/spatial differences in
vaccination has provided valuable information that can be used by
decision-makers in the development of public health policies that are
focused on improving vaccination within specific administrative or
geographical areas. However, spatial analyses of vaccination have been
very limited in Canada, as existing studies have been focused only at
certain cities (such as
Toronto\textsuperscript{\protect\hyperlink{ref-choi2021}{28}}, or
Montreal\textsuperscript{\protect\hyperlink{ref-mckinnon2021}{29}}), or
have explored differences at a province-wide
level\textsuperscript{\protect\hyperlink{ref-guay2022}{19}}. However,
there is a lack of studies that explore spatial heterogeneities in
vaccination at the intra-provincial level.

Ontario is the most populated province in Canada, and has recently seen
major changes on its healthcare structure as between 2006 and 2019, it
provided healthcare access using 14 intra-provincial divisions called
the Local Health Integrated Networks (LHINs). However, this approach was
complex and caused systemic
inequalities\textsuperscript{\protect\hyperlink{ref-tsasis2012}{30}}. In
late 2019, the 14 LHINs were phased out and the areas they covered were
incorporated into 6 Health Regions (North East, North West, Central,
Toronto, West, and East) in an effort to improve the healthcare system
of the province\textsuperscript{\protect\hyperlink{ref-dong2022}{31}}.
Because the adoption of the Health Regions has been recent, there is an
ongoing need to analyze the impact of their adoption and intra-regional
differences that might exist, which is specially important in the
context of the COVID-19 pandemic.

To address this need, in this study we analyzed differences in
self-reported COVID-19 vaccination status using socio-economic (e.g.,
income, racial/ethnic identification), and spatial information at the
level of the Health Regions, to identify the existence of
intra-provincial differences that might need to be addressed to ensure
that the healthcare system of Ontario is more inclusive and that
responds to the needs of its most vulnerable population.

@@

\hypertarget{sec-data}{%
\subsection{Data}\label{sec-data}}

We used data from the \emph{Survey of COVID-19 related Behaviours and
Attitudes}, a repeated cross sectional survey focused on the Canadian
province of Ontario that was commissioned by the Fields Institute for
Research in Mathematical Sciences (henceforth Fields) and the
Mathematical Modelling of COVID-19 Task Force under ethical guidance
from the University of Toronto. The survey collected socio-economic
information from participants (Table~\ref{tbl-covariates}), recorded
their location (using the nearest municipality), and asked information
on vaccination status by using the question ``Have you received the
first dose of the COVID vaccine?'', with possible answers ``yes'' and
``no''. The original dataset contained 39,029 entries (where each entry
corresponded to a unique respondent).

This dataset was cleaned to remove outliers that were identified during
preliminary analyses, and processing the geographical information in the
survey (city where the survey was responded) in order to match each city
to its correspondent Health Region.

\hypertarget{tbl-covariates}{}
\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.3014}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.6986}}@{}}
\caption{\label{tbl-covariates}Socio-economic factors from the Fields
COVID-19 survey}\tabularnewline
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
Variable
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Levels
\end{minipage} \\
\midrule()
\endfirsthead
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
Variable
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Levels
\end{minipage} \\
\midrule()
\endhead
Age group & 16-34,35-54,55 and over \\
Income bracket (CAD) & under 25,000, 25,000-59,999, 60,000 and above \\
Race/ethnicity & Arab/Middle Eastern, Black, East Asian/Pacific
Islander, Indigenous, Latin American, Mixed, South Asian, White
Caucasian, Other \\
\bottomrule()
\end{longtable}

The clean dataset contained responses from more than 200 different
municipalities within Ontario (Figure~\ref{fig-map}). Because of the
lack of a publicly available list of all municipalities within each
Health Region, we used a dataset of long-term care homes and LHINs to
match each city to LHIN, followed by matching each LHIN to a Health
Region following the information provided on the Ontario Health Website,
where the list of LHINs and corresponding Health Regions is available.
In the case of municipalities that did not appear in the long-term care
home dataset, we manually searched each city in the LHINs websites in
order to provide geographical information. The original dataset, clean
dataset, and details on the data cleaning process are described in
detail in the GitHub repository for this paper, which can be found at
\url{https://github.com/aimundo/Fields_COVID-19/}.

Following an assessment of the number of entries corresponding to each
Health Region in the final dataset, only 107 observations (4.3\% of the
total) corresponded to cities located in the North West and North East
Health Regions. The low representation of these Health Regions in the
dataset is noticeable in Figure~\ref{fig-map}, where the number of
cities in each Health Region is relatively low compared to the most
populated Health Regions, such as the Toronto or Central Regions. We
omitted the North East and North East Health Regions from further
analyses due to the low number of entries. Therefore, the total number
of unique entries used for analysis was 3,551 which included the East,
Central, Toronto, and West Health Regions.

\begin{figure}

{\centering \includegraphics{../data/map_data/map.pdf}

}

\caption{\label{fig-map}Geographic representation of the data collected
by the \emph{Survey of COVID-19 related Behaviours and Attitudes},
collected by the Fields Institute in Ontario. The municipalities
(cities) from where survey participants provided answers (in the clean
dataset) appear as points. The Health six Regions are color-coded.
Internal boundaries within certain Health Regions indicate areas that
belonged to the Local Integrated Health Networks (LHINs), the geographic
areas for healthcare in Ontario before the adoption of the Health
Regions.}

\end{figure}

@@

We used a logistic regression model to estimate the probability of
vaccination depending on the socio-economic factors described in
Table~\ref{tbl-covariates}, the Health Regions from Ontario indicated in
Section~\ref{sec-data}, and the interactions between Race and Health
Region, and Race and income, as previous studies have shown that
socio-economic factors and their interactions are significant predictors
of intent of vaccination and vaccination
status\textsuperscript{\protect\hyperlink{ref-nguyen2022}{32}--\protect\hyperlink{ref-cnat2022a}{34}}.

The model was fitted first to the clean dataset to obtain uncorrected
estimates. Additionally, because we identified differences between the
proportions of all the socio-economic factors included in the analysis
(Table~\ref{tbl-covariates}) and the Census data for Ontario, we used an
iterative proportional fitting procedure
(\emph{raking})\textsuperscript{\protect\hyperlink{ref-deming1940}{35}}
to correct the data using Census socio-economic data and Health Region
population totals, in order to obtain corrected estimates from the
model. Details regarding the correction can be found in the Appendix.
All analyses were conducted in R 4.2.2 using the packages
\texttt{survey}\textsuperscript{\protect\hyperlink{ref-lumley2011}{36}},\texttt{tidyverse}\textsuperscript{\protect\hyperlink{ref-wickham2019}{37}},
and \texttt{quarto}\textsuperscript{\protect\hyperlink{ref-quarto}{38}}.

@@

Table~\ref{tbl-model} shows the results of the logistic regression
models (for the uncorrected and corrected data) on vaccination status
using socio-economic factors (age group, income, race), geographical
areas (Health Regions) and the interactions between income and race and
Health Region and race. There were no statistically significant
differences in vaccination rates within the age groups from the survey,
but significant odds ratios were estimated for other covariates. Within
household income brackets, individuals with an income under CAD 25,000
or between CAD 25,000-59,999 had significantly lower odds of vaccination
than those with an income above CAD 60,000 (ORs=0.37 and 0.59,
\emph{p}=0.011 and \textless0.001, respectively). Within Race/Ethnicity,
individuals who identified as Arab/Middle Eastern, Black, or Latin
American, had significantly lower odds of vaccination than those in the
White/Caucasian group (ORs=0.31, 0.32, 0.28, and
\emph{p}=0.004,\textless0.001 and 0.004, respectively); additionally,
those individuals in the Other Race/Ethnicity group (a group that
included Southeast Asian, Filipino, West Asian, and Minorities Not
Identified Elsewhere) had even lower odds of vaccination than the other
minority groups (OR=0.22, \emph{p}\textless0.001). Regarding Health
Regions, individuals that reported living in the West Health Region
(which comprises the regions of Waterloo and Niagara, the counties of
Wellington, Essex, and Lambton, and the cities of Hamilton, Haldimand,
Brant, and Chatham-Kent) had significantly higher odds of vaccination
than those in the Health Region of Toronto (OR=1.55, \emph{p}=0.029).

@@

\hypertarget{tbl-model}{}
\begin{longtable}{lcccccc}
\caption{\label{tbl-model}Multiple Regression Analysis-Predictors of Vaccination Status }\tabularnewline

\toprule
\multicolumn{1}{c}{ } & \multicolumn{3}{c}{\textbf{Uncorrected}} & \multicolumn{3}{c}{\textbf{Corrected}} \\
\cmidrule(l{3pt}r{3pt}){2-4} \cmidrule(l{3pt}r{3pt}){5-7}
\textbf{Characteristic} & \textbf{OR} & \textbf{95\% CI} & \textbf{p-value} & \textbf{OR} & \textbf{95\% CI} & \textbf{p-value}\\
\midrule
\endfirsthead
\multicolumn{7}{@{}l}{\textit{(continued)}}\\
\toprule
\multicolumn{1}{c}{ } & \multicolumn{3}{c}{\textbf{Uncorrected}} & \multicolumn{3}{c}{\textbf{Corrected}} \\
\cmidrule(l{3pt}r{3pt}){2-4} \cmidrule(l{3pt}r{3pt}){5-7}
\textbf{Characteristic} & \textbf{OR} & \textbf{95\% CI} & \textbf{p-value} & \textbf{OR} & \textbf{95\% CI} & \textbf{p-value}\\
\midrule
\endhead

\endfoot
\bottomrule
\endlastfoot
\textbf{Age Group} &  &  &  &  &  & \\
\hspace{1em}16\_34 & — & — &  & — & — & \\
\hspace{1em}35\_54 & 0.93 & 0.77, 1.13 & 0.5 & 0.90 & 0.67, 1.21 & 0.5\\
\hspace{1em}55\_and\_over & 0.74 & 0.61, 0.89 & 0.002 & 0.99 & 0.74, 1.32 & >0.9\\
\textbf{Income} &  &  &  &  &  & \\
\hspace{1em}60000\_and\_above & — & — &  & — & — & \\
\hspace{1em}25000\_59999 & 0.59 & 0.41, 0.84 & 0.004 & 0.59 & 0.39, 0.89 & 0.011\\
\hspace{1em}under\_25000 & 0.41 & 0.29, 0.58 & <0.001 & 0.37 & 0.25, 0.56 & <0.001\\
\textbf{Race} &  &  &  &  &  & \\
\hspace{1em}white\_caucasian & — & — &  & — & — & \\
\hspace{1em}arab\_middle\_eastern & 0.24 & 0.12, 0.50 & <0.001 & 0.31 & 0.14, 0.69 & 0.004\\
\hspace{1em}black & 0.30 & 0.17, 0.54 & <0.001 & 0.32 & 0.17, 0.60 & <0.001\\
